---
title: "GyNet Choix Design - étude des données avant DESeq2"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Gynet/"
knitr::opts_knit$set(root.dir = path)

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}
# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("DEGreport")
library("apeglm")
library("ggrepel")
library("GGally")
library("gridExtra")
library("mixOmics")

library("dplyr")
library("ggplot2")
library("factoextra")
library("reshape2")
library("tidyr")
library("splines")
library(viridis)
library(conflicted)
# install.packages("estimate", repos="http://R-Forge.R-project.org")

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Objectif

Etudier la QC sur les échantillons GyNet, avant recherche de gènes différentiellement exprimés via DESeq2.<br/>

<br/>
<br/>

Analyse combinée des 168 samples, puis en séparant par indication : Endo (101 samples) et Cervix (67 samples).<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./coldata_GyNet.csv', sep=",")
coldata$BMI <- coldata$Weight / (coldata$Height)^2
coldata$HR <- ifelse(
  coldata$Estrogen_endo == "0" & coldata$Progesterone_endo == "0",
  "neg",
  "pos"
)
coldata$HR_2 <- case_when(
  coldata$Estrogen_endo == "0" ~ "E_neg",
  coldata$Estrogen_endo == "1" ~ "E_pos",
  coldata$Progesterone_endo == "0" ~ "P_neg",
  coldata$Progesterone_endo == "1" ~ "P_pos",
)
# coldata <- subset( coldata, select = - c(xx, xx))

# scale and center the numeric columns
numeric_columns <- c('Age', 'Weight', 'Height', 'BMI', 'OS', 'PFS', 'Cycles_NP137', 'Cycles_carboplatin', 'Cycles_paclitaxel', 'Cycles_pembrolizumab', 'Purity', 'CT_perc', 'QC_duplication_perc', 'QC_alignedM_Salmon', 'QC_uniq_aligned')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
# convert to factors the non-numeric columns
factor_columns <- colnames(coldata) %!in% numeric_columns 
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)


# countData (gene counts, not transcript's)
counts_genes <- read.table("./results_rnaseq_NFcore/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
#counts_genes <- as.matrix(counts_genes)

# keeping only the protein-coding genes, as we used a protocol capturing only exons (???)
gtf <- import('/home/eugenie-modolo/Documents/Reference_files/gencode.v49.chr_patch_hapl_scaff.annotation.gtf')
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[which(rownames(counts_genes) %in% keep_genes), ]

count_matrix <- as.matrix(count_matrix_pcg[, count_columns])

coldata <- data.frame(coldata, row.names = count_columns)

coldata_wo_rep <- subset( coldata, select = -c(Age, Height, Weight, BMI, OS, PFS, Cycles_NP137, Cycles_carboplatin, Cycles_paclitaxel, Cycles_pembrolizumab, Purity, CT_perc, QC_duplication_perc, QC_alignedM_Salmon, QC_uniq_aligned) )

for (col in colnames(coldata_wo_rep)) {
    if (endsWith(col, '_cr')) {
        coldata_wo_rep[[col]] <- as.numeric(as.vector(coldata_wo_rep[[col]]))
    }
}

# creating the Endo and Cervix data
coldata_Endo <- coldata_wo_rep[which(coldata_wo_rep$Indication == "Endo"),]
coldata_Endo <- subset( coldata_Endo, select = -c(Indication, HPV_all, HPV16) ) # colonne qui ne prend qu'une valeur et colonnes qui n'ont que des NA
coldata_Endo <- droplevels.data.frame(coldata_Endo)
count_matrix_Endo <- count_matrix[, which(coldata_wo_rep$Indication == "Endo")]

coldata_Cervix <- coldata_wo_rep[which(coldata_wo_rep$Indication == "Cervix"),]
coldata_Cervix <- subset( coldata_Cervix, select = -c(Indication, Mut_Status_endo, Estrogen_endo, Progesterone_endo, HR, HR_2) ) # colonne qui ne prend qu'une valeur et colonnes qui n'ont que des NA
coldata_Cervix <- droplevels.data.frame(coldata_Cervix)
count_matrix_Cervix <- count_matrix[, which(coldata_wo_rep$Indication == "Cervix")]

# removing the missing/problematic data for LRT
coldata_Endo_Primary <- coldata_Endo[which(coldata_Endo$Site == "Primary" & !is.na(coldata_Endo$Estrogen_endo) & !is.na(coldata_Endo$Progesterone_endo) & !is.na(coldata_Endo$best_response)),]
count_matrix_Endo_Primary <- count_matrix_Endo[, which(coldata_Endo$Site == "Primary" & !is.na(coldata_Endo$Estrogen_endo) & !is.na(coldata_Endo$Progesterone_endo) & !is.na(coldata_Endo$best_response))]
coldata_Cervix_Primary <- coldata_Cervix[which(coldata_Cervix$Site == "Primary" & !is.na(coldata_Cervix$HPV_all) & !is.na(coldata_Cervix$HPV16)),]
count_matrix_Cervix_Primary <- count_matrix_Cervix[, which(coldata_Cervix$Site == "Primary" & !is.na(coldata_Cervix$HPV_all) & !is.na(coldata_Cervix$HPV16))] 

# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

```

Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix)))` gène(s) au nom dupliqué, sur un total de `r nrow(count_matrix)` protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(coldata) == colnames(count_matrix)) == length(rownames(coldata))`.<br/>


## Utilisation du LRT pour choisir le design : Endo


```{r forward selection strategy LRT Endo}

# Function to check if a design matrix is full rank
check_design_rank <- function(count_matrix, coldata, formula) {
  
  tryCatch({
    # Create model matrix
    mm <- model.matrix(formula, data = coldata)
    # Check rank
    is_full_rank <- qr(mm)$rank == ncol(mm)
    
    return(list(
      is_valid = is_full_rank,
      rank = qr(mm)$rank,
      ncol = ncol(mm)
    ))
  }, error = function(e) {
    return(list(
      is_valid = FALSE,
      error = e$message
    ))
  })
}

# Sequential forward selection using LRT
test_variable_influence_lrt_sequential <- function(count_matrix,
                                                    coldata,
                                                    variables = NULL,
                                                    nber_round = NULL,
                                                    padj_cutoff = 0.05,
                                                    lfc_cutoff = 0.58) {
  
  # Get all variables from coldata if not specified
  if (is.null(variables)) {
    variables <- colnames(coldata)
  } else {
    missing_vars <- setdiff(variables, colnames(coldata))
    if (length(missing_vars) > 0) {
      stop("These variables not found in coldata: ", paste(missing_vars, collapse = ", "))
    }
  }
  if (is.null(nber_round)) {
    last_round <- 0
  } else {
    last_round <- length(variables) - nber_round
  }
  
  # Storage for all results
  all_rounds_results <- list()
  selected_variables <- character()
  remaining_variables <- variables
  
  round_num <- 1
  
  # Continue until all variables are tested
  while (length(remaining_variables) > last_round) {
    
    # cat("\n========================================\n")
    # cat("ROUND", round_num, "\n")
    # cat("========================================\n")
    
    if (round_num == 1) {
      # cat("Testing variables alone (reduced = ~1)\n\n")
      reduced_formula <- "~1"
    } else {
      # cat("Current design:", paste("~", paste(selected_variables, collapse = " + ")), "\n")
      # cat("Testing addition of remaining variables\n\n")
      reduced_formula <- paste("~", paste(selected_variables, collapse = " + "))
    }
    
    # Storage for this round
    round_results <- list()
    skipped_tests <- list()
    
    # Test each remaining variable
    for (var in remaining_variables) {
      
      if (round_num == 1) {
        full_formula_str <- paste("~", var)
        # cat("Testing:", var, "vs ~1\n")
      } else {
        full_formula_str <- paste("~", paste(c(var, selected_variables), collapse = " + "))
        # cat("Testing:", var, "added to current design\n")
      }
      
      # Check design validity
      full_formula <- as.formula(full_formula_str)
      rank_check <- check_design_rank(count_matrix, coldata, full_formula)
      
      if (!rank_check$is_valid) {
        # cat("  SKIPPED: Design matrix not full rank\n")
        skipped_tests[[var]] <- "Not full rank"
        next
      }
      
      tryCatch({
        # Create dds
        dds <- DESeqDataSetFromMatrix(
          countData = count_matrix,
          colData = coldata,
          design = full_formula
        )
        
        # Run LRT
        dds_lrt <- DESeq(dds, test = "LRT", 
                        reduced = as.formula(reduced_formula), 
                        quiet = TRUE)
        res <- results(dds_lrt)
        
        # Count DEGs
        n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
        n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                         na.rm = TRUE)
        
        round_results[[var]] <- data.frame(
          round = round_num,
          tested_variable = var,
          current_design = ifelse(round_num == 1, "none", 
                                 paste(selected_variables, collapse = " + ")),
          full_model = full_formula_str,
          reduced_model = reduced_formula,
          n_deg_padj_only = n_deg_padj,
          n_deg_padj_and_lfc = n_deg_both,
          stringsAsFactors = FALSE
        )
        
        # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
        
      }, error = function(e) {
        # cat("  ERROR:", e$message, "\n")
        skipped_tests[[var]] <- e$message
      })
    }
    
    # Combine results for this round
    if (length(round_results) == 0) {
      # cat("\nNo successful tests in this round. Stopping.\n")
      break
    }
    
    round_df <- do.call(rbind, round_results)
    rownames(round_df) <- NULL
    
    # Select best variable (most DEGs with padj + LFC)
    best_idx <- which.max(round_df$n_deg_padj_and_lfc)
    best_variable <- round_df$tested_variable[best_idx]
    best_n_deg <- round_df$n_deg_padj_and_lfc[best_idx]
    
    # cat("\n--- ROUND", round_num, "SUMMARY ---\n")
    # cat("Best variable:", best_variable, "with", best_n_deg, "DEGs\n")
    
    # Add to results
    all_rounds_results[[paste0("round_", round_num)]] <- round_df
    
    # Update for next round
    selected_variables <- c(best_variable, selected_variables)
    remaining_variables <- setdiff(remaining_variables, best_variable)
    
    round_num <- round_num + 1
  }
  
  # Combine all rounds
  final_results <- do.call(rbind, all_rounds_results)
  rownames(final_results) <- NULL
  
  # Summary
  # cat("\n========================================\n")
  # cat("FINAL SUMMARY\n")
  # cat("========================================\n")
  # cat("Variables selected in order:\n")
  for (i in seq_along(selected_variables)) {
    var_info <- final_results[final_results$round == i & 
                              final_results$tested_variable == rev(selected_variables)[i], ]
    # cat(i, ".", rev(selected_variables)[i], "- DEGs:", 
    #     var_info$n_deg_padj_and_lfc, "\n")
  }
  
  # cat("\nFinal design:", paste("~", paste(selected_variables, collapse = " + ")), "\n")
  
  return(list(
    results = final_results,
    selected_order = rev(selected_variables),
    selection_summary = data.frame(
      rank = 1:length(selected_variables),
      variable = rev(selected_variables),
      n_deg = sapply(rev(selected_variables), function(v) {
        final_results[final_results$tested_variable == v, "n_deg_padj_and_lfc"][1]
      })
    )
  ))
}



# Heatmap showing DEGs for each variable at each selection round
plot_sequential_selection_heatmap <- function(lrt_results, criterion = "padj_and_lfc") {
  
  results_df <- lrt_results$results
  
  # Choose which DEG count to use
  if (criterion == "padj_only") {
    results_df$n_deg <- results_df$n_deg_padj_only
    legend_title <- "DEGs (padj only)"
  } else {
    results_df$n_deg <- results_df$n_deg_padj_and_lfc
    legend_title <- "DEGs (padj + LFC)"
  }
  
  # Order variables by selection order
  variable_order <- lrt_results$selected_order
  if (length(variable_order) < length(unique(results_df$tested_variable))) {
    for (var in unique(results_df$tested_variable)) {
      if (var %!in% variable_order) {
        variable_order <- c(variable_order, var)
      }
    }
  }
  results_df$tested_variable <- factor(results_df$tested_variable, 
                                       levels = variable_order)
  
  # Create the plot
  p <- ggplot(results_df, aes(x = factor(round), y = tested_variable, fill = n_deg)) +
    geom_tile(color = "white", size = 1) +
    geom_text(aes(label = n_deg), color = "white", size = 4, fontface = "bold") +
    scale_fill_viridis(option = "plasma", name = legend_title,
                      na.value = "grey90") +
    labs(title = "Sequential Variable Selection: DEGs per Round",
         subtitle = "Variables ordered by selection rank (down = selected first)",
         x = "Selection Round",
         y = "Variable") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 11),
          axis.text.y = element_text(size = 10),
          plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, size = 10),
          legend.position = "right",
          panel.grid = element_blank())
  
  return(p)
}


# Summary barplot showing final selection order # ne fonctionne pas : donne le nombre de DEGs du premier round
plot_selection_summary <- function(lrt_results, criterion = "padj_and_lfc") {
  
  summary_df <- lrt_results$selection_summary
  
  # Order by rank
  summary_df$variable <- factor(summary_df$variable, levels = summary_df$variable)
  
  p <- ggplot(summary_df, aes(x = variable, y = n_deg, fill = rank)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = n_deg), vjust = -0.5, size = 4) +
    scale_fill_viridis(option = "viridis", direction = -1, 
                      name = "Selection\nRank") +
    labs(title = "Variable Selection Order and Impact",
         subtitle = "Number of DEGs when variable was selected",
         x = "Variable (ordered by selection)",
         y = "Number of DEGs") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
          plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, size = 10))
  
  return(p)
}

# Combined visualization
plot_sequential_results <- function(lrt_results, criterion = "padj_and_lfc") {
  
  p1 <- plot_sequential_selection_heatmap(lrt_results, criterion)
  p2 <- plot_selection_summary(lrt_results, criterion)
  
  grid.arrange(p1, p2, nrow = 2, heights = c(2, 1))
}

# The heatmap will show:
# - Rows: Variables ordered by when they were selected (top = selected first)
# - Columns: Selection rounds (1, 2, 3, ...)
# - Color intensity: Number of DEGs found when that variable was tested in that round
# - Diagonal pattern: The selected variable in each round (brightest in its selection round)
# The barplot shows the final selection order and how many DEGs each variable contributed when it was added to the design.


# # Run sequential selection
# sequential_results <- test_variable_influence_lrt_sequential(
#   count_matrix = count_matrix_Endo_Primary,
#   coldata = coldata_Endo_Primary,
#   variables = c("Arm", "Batch", "Death_status", "Prog_status", "best_response", "Nber_Prior_chemo", "ecog", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "OS_cr", "PFS_cr", "Cycles_NP137_cr", "Cycles_carboplatin_cr", "Cycles_paclitaxel_cr", "Cycles_pembrolizumab_cr", "Purity_cr", "CT_perc_cr", "HR", "HR_2", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr"),  # or NULL for all
#   nber_round = 5, # or NULL for all rounds
#   padj_cutoff = 0.05,
#   lfc_cutoff = 0.58
# )
# 
# # View results table
# # print(sequential_results$results)
# 
# # View selection order
# print(sequential_results$selection_summary)
# 
# # Visualize
# plot_sequential_selection_heatmap(sequential_results, criterion = "padj_and_lfc")
# # plot_selection_summary(sequential_results, criterion = "padj_and_lfc")
# 
# # Combined plot
# # plot_sequential_results(sequential_results, criterion = "padj_and_lfc")
# 
# # # Save results
# # write.csv(sequential_results$results, "sequential_lrt_results.csv", row.names = FALSE)
# # write.csv(sequential_results$selection_summary, "variable_selection_order.csv", row.names = FALSE)

```


```{r forward selection strategy LRT with duplicates Endo}

# Modification of test_variable_influence_lrt_sequential using the information of linked variables
test_variable_influence_lrt_sequential_duplicates <- function(count_matrix,
                                                   coldata,
                                                   duplicates,
                                                   variables = NULL,
                                                   padj_cutoff = 0.05,
                                                   lfc_cutoff = 0.58) {
  
  # Get all variables from coldata if not specified
  if (is.null(variables)) {
    variables <- colnames(coldata)
  } else {
    missing_vars <- setdiff(variables, colnames(coldata))
    if (length(missing_vars) > 0) {
      stop("These variables not found in coldata: ", paste(missing_vars, collapse = ", "))
    }
  }
  duplicates_list <- do.call(c, duplicates)

  # Storage for all results
  all_rounds_results <- list()
  selected_variables <- character()
  remaining_variables <- variables
  
  round_num <- 1
  check_zero_deg <- 1
  
  # Continue until all variables are tested
  while (length(remaining_variables) > 0 & check_zero_deg == 1) {
    
    # cat("\n========================================\n")
    # cat("ROUND", round_num, "\n")
    # cat("========================================\n")
    
    if (round_num == 1) {
      # cat("Testing variables alone (reduced = ~1)\n\n")
      reduced_formula <- "~1"
    } else {
      # cat("Current design:", paste("~", paste(selected_variables, collapse = " + ")), "\n")
      # cat("Testing addition of remaining variables\n\n")
      reduced_formula <- paste("~", paste(selected_variables, collapse = " + "))
    }
    
    # Storage for this round
    round_results <- list()
    skipped_tests <- list()
    
    # Test each remaining variable
    for (var in remaining_variables) {
      
      if (round_num == 1) {
        full_formula_str <- paste("~", var)
        # cat("Testing:", var, "vs ~1\n")
      } else {
        full_formula_str <- paste("~", paste(c(var, selected_variables), collapse = " + "))
        # cat("Testing:", var, "added to current design\n")
      }
      
      # Check design validity
      full_formula <- as.formula(full_formula_str)
      rank_check <- check_design_rank(count_matrix, coldata, full_formula)
      
      if (!rank_check$is_valid) {
        # cat("  SKIPPED: Design matrix not full rank\n")
        skipped_tests[[var]] <- "Not full rank"
        next
      }
      
      tryCatch({
        # Create dds
        dds <- DESeqDataSetFromMatrix(
          countData = count_matrix,
          colData = coldata,
          design = full_formula
        )
        
        # Run LRT
        dds_lrt <- DESeq(dds, test = "LRT", 
                         reduced = as.formula(reduced_formula), 
                         quiet = TRUE)
        res <- results(dds_lrt)
        
        # Count DEGs
        n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
        n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                          na.rm = TRUE)
        
        round_results[[var]] <- data.frame(
          round = round_num,
          tested_variable = var,
          current_design = ifelse(round_num == 1, "none", 
                                  paste(selected_variables, collapse = " + ")),
          full_model = full_formula_str,
          reduced_model = reduced_formula,
          n_deg_padj_only = n_deg_padj,
          n_deg_padj_and_lfc = n_deg_both,
          stringsAsFactors = FALSE
        )
        
        # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
        
      }, error = function(e) {
        # cat("  ERROR:", e$message, "\n")
        skipped_tests[[var]] <- e$message
      })
    }
    
    # Combine results for this round
    if (length(round_results) == 0) {
      # cat("\nNo successful tests in this round. Stopping.\n")
      break
    }
    
    round_df <- do.call(rbind, round_results)
    rownames(round_df) <- NULL
    
    # Select best variable (most DEGs with padj + LFC)
    best_idx <- order(round_df$n_deg_padj_and_lfc, round_df$n_deg_padj_only, 
                  decreasing = TRUE)[1] # which.max(round_df$n_deg_padj_and_lfc)
    best_variable <- round_df$tested_variable[best_idx]
    best_n_deg <- round_df$n_deg_padj_and_lfc[best_idx]
        
    if (sum(round_df$n_deg_padj_and_lfc) == 0 & sum(round_df$n_deg_padj_only) == 0) {
      # cat("\nNo variable gave at least 1 DEG with the LRT method. Stopping.\n")
      check_zero_deg <- 0
    }
    
    # cat("\n--- ROUND", round_num, "SUMMARY ---\n")
    # cat("Best variable:", best_variable, "with", best_n_deg, "DEGs\n")
    
    # Add to results
    all_rounds_results[[paste0("round_", round_num)]] <- round_df
    
    # Update for next round
    selected_variables <- c(best_variable, selected_variables)
    if (best_variable %in% duplicates_list) {
      which_sublist <- which(sapply(duplicates, function(x) best_variable %in% x))
      remaining_variables <- setdiff(remaining_variables, duplicates[[which_sublist]])
    }
    else {
      remaining_variables <- setdiff(remaining_variables, best_variable)
    }
    
    round_num <- round_num + 1
  }
  
  # Combine all rounds
  final_results <- do.call(rbind, all_rounds_results)
  rownames(final_results) <- NULL
  
  # Summary
  # cat("\n========================================\n")
  # cat("FINAL SUMMARY\n")
  # cat("========================================\n")
  # cat("Variables selected in order:\n")
  for (i in seq_along(selected_variables)) {
    var_info <- final_results[final_results$round == i & 
                                final_results$tested_variable == rev(selected_variables)[i], ]
    # cat(i, ".", rev(selected_variables)[i], "- DEGs:", 
    #     var_info$n_deg_padj_and_lfc, "\n")
  }
  
  # cat("\nFinal design:", paste("~", paste(selected_variables, collapse = " + ")), "\n")
  
  return(list(
    results = final_results,
    selected_order = rev(selected_variables),
    selection_summary = data.frame(
      rank = 1:length(selected_variables),
      variable = rev(selected_variables),
      n_deg = sapply(rev(selected_variables), function(v) {
        final_results[final_results$tested_variable == v, "n_deg_padj_and_lfc"][1]
      })
    )
  ))
}


duplicates <- list()
duplicates$QC <- c("QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr")
duplicates$purity <- c("Purity_cr", "CT_perc_cr")
duplicates$HR <- c("HR", "HR_2")
duplicates$var_interest <- c("OS_cr", "PFS_cr")
duplicates$clinical_patient <- c("ecog", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr")
duplicates$clinical_ante <- c("Cycles_NP137_cr", "Cycles_carboplatin_cr", "Cycles_paclitaxel_cr", "Cycles_pembrolizumab_cr")
duplicates$clinical_resp <- c("Death_status", "Prog_status", "best_response")

# Run sequential selection
sequential_results <- test_variable_influence_lrt_sequential_duplicates(
  count_matrix = count_matrix_Endo_Primary,
  coldata = coldata_Endo_Primary,
  duplicates = duplicates,
  variables = c("Arm", "Batch", "Death_status", "Prog_status", "best_response", "Nber_Prior_chemo", "ecog", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "OS_cr", "PFS_cr", "Cycles_NP137_cr", "Cycles_carboplatin_cr", "Cycles_paclitaxel_cr", "Cycles_pembrolizumab_cr", "Purity_cr", "CT_perc_cr", "HR", "HR_2", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr"),  # or NULL for all
  padj_cutoff = 0.05,
  lfc_cutoff = 0.58
)

# View results table
# print(sequential_results$results)

# View selection order
print(sequential_results$selection_summary)

# Visualize
plot_sequential_selection_heatmap(sequential_results, criterion = "padj_and_lfc")

```


## Utilisation du LRT pour choisir le design : Cervix


```{r forward selection strategy LRT Cervix}

duplicates$HPV <- c("HPV16", "HPV_all")

# Run sequential selection
# sequential_results <- test_variable_influence_lrt_sequential(
#   count_matrix = count_matrix_Cervix_Primary,
#   coldata = coldata_Cervix_Primary,
#   variables = c("Arm", "Batch", "Death_status", "Prog_status", "best_response", "Nber_Prior_chemo", "ecog", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "OS_cr", "PFS_cr", "Cycles_NP137_cr", "Cycles_carboplatin_cr", "Cycles_paclitaxel_cr", "Cycles_pembrolizumab_cr", "Purity_cr", "CT_perc_cr", "HPV16", "HPV_all", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr"),  # or NULL for all
#   nber_round = 5, # or NULL for all rounds
#   padj_cutoff = 0.05,
#   lfc_cutoff = 0.58
# )
sequential_results <- test_variable_influence_lrt_sequential_duplicates(
  count_matrix = count_matrix_Cervix_Primary,
  coldata = coldata_Cervix_Primary,
  duplicates = duplicates,
  variables = c("Arm", "Batch", "Death_status", "Prog_status", "best_response", "Nber_Prior_chemo", "ecog", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "OS_cr", "PFS_cr", "Cycles_NP137_cr", "Cycles_carboplatin_cr", "Cycles_paclitaxel_cr", "Cycles_pembrolizumab_cr", "Purity_cr", "CT_perc_cr", "HPV16", "HPV_all", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr"),  # or NULL for all
  padj_cutoff = 0.05,
  lfc_cutoff = 0.58
)

# View results table
# print(sequential_results$results)

# View selection order
print(sequential_results$selection_summary)

# Visualize
plot_sequential_selection_heatmap(sequential_results, criterion = "padj_and_lfc")

```


# Conclusion et remarques (à modifier !!)

Choix du design, selon l'indication :<br/>
- variables à inclure pour retirer les effets non recherchés (on en choisit une par catégorie pour qu'elles ne risquent pas d'être orthogonales) : <br/>
  - pour les données cliniques : Age_cr est le mieux placé dans 2 des trois (LRT, PLS_600). Il pourrait être intéressant de tester en plus (si non-orthogonal) : (i) Sex (bien classé en LRT et PLS_all), (ii) Height_cr (bien classé en LRT, mais pas en PLS et pas de logique biologique ...), (iii) les 4 variables de survie citées ci-dessous <br/> 
  - pour les données de QC : QC_aligned_M_cr est le mieux placé dans 2 des trois (LRT, PLS_all) <br/> 
  - pour l'étude des pré : il faudra peut-être ajouter Time dans les variables avant celle d'intérêt, car il est très haut dans LRT mais pas les PLS.<br/>
<br/>
- variables explicatives :<br/>
  - pour la recherche de biomarqueurs de survie (étude uniquement des pre) : il y a le choix entre : (i) best_lesion_perc_cr (très bon en valeur unique dans le LRT et dans PLS_600), (ii) Best_response_2 (bien classé dans les 3), (iii) tumor_spline_slope_cr (bon en LRT mais pas super en PLS), (iv) Prog_status (pas mal classé dans les 3).<br/>
  - pour la différence pre/post : Time ressort très nettement en LRT (c'était la seule variable pour cette condition de toutes façons).<br/>


# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
