---
title: "GyNet QC - étude des données avant DESeq2"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Gynet/"
# setwd(path)
knitr::opts_knit$set(root.dir = path)

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}
# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("DEGreport")
library("apeglm")
library("ggrepel")
library("GGally")
library("gridExtra")
library("mixOmics")

library("dplyr")
library("ggplot2")
library("factoextra")
library("reshape2")
library("tidyr")
library("splines")
library(reshape2)
library(estimate)
library(tibble)
library(txdbmaker)

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Objectif

Etudier la QC sur les échantillons GyNet, avant recherche de gènes différentiellement exprimés via DESeq2.<br/>
<br/>

Analyse combinée des 168 samples, puis en séparant par indication : Endo (101 samples) et Cervix (67 samples).<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./coldata_GyNet.csv', sep=",")
coldata$BMI <- coldata$Weight / (coldata$Height)^2
coldata$HR <- ifelse(
  coldata$Estrogen_endo == "0" & coldata$Progesterone_endo == "0",
  "neg",
  "pos"
)
coldata$HR_2 <- case_when(
  coldata$Estrogen_endo == "0" ~ "E_neg",
  coldata$Estrogen_endo == "1" ~ "E_pos",
  coldata$Progesterone_endo == "0" ~ "P_neg",
  coldata$Progesterone_endo == "1" ~ "P_pos",
)
# coldata <- subset( coldata, select = - c(xx, xx))

# scale and center the numeric columns
numeric_columns <- c('Age', 'Weight', 'Height', 'BMI', 'OS', 'PFS', 'Cycles_NP137', 'Cycles_carboplatin', 'Cycles_paclitaxel', 'Cycles_pembrolizumab', 'Purity', 'CT_perc', 'QC_duplication_perc', 'QC_alignedM_Salmon', 'QC_uniq_aligned')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
# convert to factors the non-numeric columns
factor_columns <- colnames(coldata) %!in% numeric_columns 
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)


# countData (gene counts, not transcript's)
counts_genes <- read.table("./results_rnaseq_NFcore/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
#counts_genes <- as.matrix(counts_genes)
count_matrix <- as.matrix(counts_genes[, count_columns])

# adding gene length to counts_genes
txdb <- txdbmaker::makeTxDbFromGFF("/home/eugenie-modolo/Documents/Reference_files/gencode.v49.chr_patch_hapl_scaff.annotation.gtf",format="gtf")
# collect the exons per gene id
exons.list.per.gene <- exonsBy(txdb,by="gene")
# then for each gene, reduce all the exons to a set of non overlapping exons, calculate their lengths (widths) and sum then
exonic.gene.sizes <- as.data.frame(sum(width(GenomicRanges::reduce(exons.list.per.gene))))
match_index <- match(rownames(counts_genes), rownames(exonic.gene.sizes))
counts_genes$gene_length <- exonic.gene.sizes[match_index, 1]

# keeping only the protein-coding genes, as we used a protocol capturing only exons
gtf <- import('/home/eugenie-modolo/Documents/Reference_files/gencode.v49.chr_patch_hapl_scaff.annotation.gtf')
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[which(rownames(counts_genes) %in% keep_genes), ]
count_matrix_pcg <- as.matrix(count_matrix_pcg[, count_columns])

# Working with gene symbols
count_matrix_genesymbol <- count_matrix
rownames(count_matrix_genesymbol) <- counts_genes$gene_name  # Gene names become row names
count_matrix_pcg_genesymbol <- count_matrix_pcg
rownames(count_matrix_pcg_genesymbol) <- counts_genes[match(rownames(count_matrix_pcg_genesymbol), rownames(counts_genes), nomatch = 0),'gene_name']
# duplicated gene symbols: /!\ we make the sum of their raw counts (hypothesis: they code for the same protein function)
count_matrix_genesymbol <- rowsum(count_matrix_genesymbol, rownames(count_matrix_genesymbol))
count_matrix_pcg_genesymbol <- rowsum(count_matrix_pcg_genesymbol, rownames(count_matrix_pcg_genesymbol))

coldata <- data.frame(coldata, row.names = count_columns)

```

<br/>
<br/>
Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix)))` gène(s) au nom dupliqué, sur un total de `r nrow(count_matrix)` protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(coldata) == colnames(count_matrix)) == length(rownames(coldata))`.<br/>


## Analyses sur le jeu de données global (Endo + Cervix)

```{r data overview}

# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ Arm
)

dds <- DESeq(dds)
norm.count_gynet <- as.data.frame(counts(dds, normalized=TRUE))

# checking the DESeq2 dispersion 
plotDispEsts(dds, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Arm'): Cooks' distance")

```

### Purity scores

```{r purity scores}

# expressions for GSEA: normalised and divided by length
dds_gene_symbol <- DESeqDataSetFromMatrix(
  countData = count_matrix_genesymbol,
  colData = coldata,
  design = ~ Arm
)
dds_gene_symbol <- DESeq(dds_gene_symbol)
norm.count_genesymbol <- as.data.frame(counts(dds_gene_symbol, normalized=TRUE))
# norm.count_genesymbol_K <- as.matrix(norm.count_genesymbol/counts_genes[which(counts_genes$gene_name %in% rownames(norm.count_genesymbol)),'gene_length'])
gene_lengths <- counts_genes$gene_length[match(rownames(norm.count_genesymbol), counts_genes$gene_name)]
norm.count_genesymbol_K <- as.matrix(norm.count_genesymbol/gene_lengths)

dds_pcg_gene_symbol <- DESeqDataSetFromMatrix(
  countData = count_matrix_pcg_genesymbol,
  colData = coldata,
  design = ~ Arm
)
dds_pcg_gene_symbol <- DESeq(dds_pcg_gene_symbol)
norm.count_pcg_genesymbol <- as.data.frame(counts(dds_pcg_gene_symbol, normalized=TRUE))
# norm.count_pcg_genesymbol_K <- as.matrix(norm.count_pcg_genesymbol/counts_genes[which(counts_genes$gene_name %in% rownames(norm.count_pcg_genesymbol)),'gene_length'])
gene_lengths <- counts_genes$gene_length[match(rownames(norm.count_pcg_genesymbol), counts_genes$gene_name)]
norm.count_pcg_genesymbol_K <- as.matrix(norm.count_pcg_genesymbol/gene_lengths)

# write.table(norm.count_genesymbol_K,"./scripts/Mehlen_Gynet_scripts/norm.count_genesymbol_K.tsv", row.names = T, col.names = T, sep="\t")
# write.table(norm.count_pcg_genesymbol_K,"./scripts/Mehlen_Gynet_scripts/norm.count_pcg_genesymbol_K.tsv", row.names = T, col.names = T, sep="\t")
# /!\ il faut modifier le fichier pour décaler les noms de samples et ajouter "GeneSymbol" comme nom de la première colonne

# Purity scores (package estimate)
# GyNey_Expr <- system.file("extdata", "./scripts/Mehlen_Gynet_scripts/count_matrix_genesymbol_pcg.tsv", package="estimate")
estimate::filterCommonGenes(input.f = "./scripts/Mehlen_Gynet_scripts/norm.count_pcg_genesymbol_K.tsv", output.f = "./scripts/Mehlen_Gynet_scripts/norm.count_pcg_genesymbol_K.gct", id="GeneSymbol")
estimate::estimateScore("./scripts/Mehlen_Gynet_scripts/norm.count_pcg_genesymbol_K.gct", "./scripts/Mehlen_Gynet_scripts/OV_estimate_score_pcg.gct", platform = "illumina")
# estimate::plotPurity(scores="./scripts/Mehlen_Gynet_scripts/OV_estimate_score.gct", samples="all_samples", platform="illumina") # ne fonctionne pas, on fait les plot autrement :

# Plotting Estimate scores
read_estimate_scores <- function(gct_file) {
  
  # Read GCT file (skip first 2 lines which are headers)
  gct_data <- read.delim(gct_file, skip = 2, header = TRUE, row.names = 1)
  
  # The scores are typically in rows: StromalScore, ImmuneScore, ESTIMATEScore
  # Transpose to get samples as rows
  scores_df <- as.data.frame(t(gct_data[,which(colnames(gct_data) != "Description")]))
  
  # Clean up column names if needed
  colnames(scores_df) <- gsub("\\.", "", colnames(scores_df))
  
  return(scores_df)
}

estimate_scores <- read_estimate_scores("./scripts/Mehlen_Gynet_scripts/OV_estimate_score_pcg.gct")

# Calculate tumor purity using the formula from the ESTIMATE paper: Purity = cos(0.6049872018 + 0.0001467884 * ESTIMATE score)
calculate_tumor_purity <- function(estimate_scores) {
  
  # Extract ESTIMATE score column
  if ("ESTIMATEScore" %in% colnames(estimate_scores)) {
    estimate_score <- as.numeric(estimate_scores$ESTIMATEScore)
  } else if ("ESTIMATE.score" %in% colnames(estimate_scores)) {
    estimate_score <- as.numeric(estimate_scores$ESTIMATE.score)
  } else {
    # Find column with "ESTIMATE" in name
    estimate_col <- grep("ESTIMATE", colnames(estimate_scores), ignore.case = TRUE, value = TRUE)[1]
    estimate_score <- as.numeric(estimate_scores[[estimate_col]])
  }
  
  # Calculate purity
  purity <- cos(0.6049872018 + 0.0001467884 * estimate_score)
  
  # Add to dataframe
  estimate_scores$TumorPurity <- purity
  
  return(estimate_scores)
}

# Calculate purity
estimate_scores <- calculate_tumor_purity(estimate_scores)
estimate_scores <- estimate_scores %>%
    rownames_to_column("Sample")
# vérif que les samples sont dans le même ordre
# sum(estimate_scores$Sample == rownames(coldata)) == length(estimate_scores$Sample)
estimate_scores$Cohort <- case_when(
  coldata$Indication == 'Endo' & coldata$Arm == 'A' ~ "Endo_A",
  coldata$Indication == 'Endo' & coldata$Arm == 'B' ~ "Endo_B",
  coldata$Indication == 'Endo' & coldata$Arm == 'C' ~ "Endo_C",
  coldata$Indication == 'Endo' & coldata$Arm == 'D' ~ "Endo_D",
  coldata$Indication == 'Cervix' & coldata$Arm == 'A' ~ "Cervix_A",
  coldata$Indication == 'Cervix' & coldata$Arm == 'B' ~ "Cervix_B",
  coldata$Indication == 'Cervix' & coldata$Arm == 'C' ~ "Cervix_C",
  coldata$Indication == 'Cervix' & coldata$Arm == 'D' ~ "Cervix_D"
)

# # Plots 1
# plot_estimate_scores <- function(scores_df) {
#   
#   # Prepare data for plotting
#   scores_long <- scores_df %>%
#     # rownames_to_column("Sample") %>%
#     # select(Sample, StromalScore, ImmuneScore, ESTIMATEScore, TumorPurity) %>%
#     pivot_longer(cols = -Sample, names_to = "Score_Type", values_to = "Score")
#   
#   # Separate purity from other scores (different scale)
#   scores_plot <- scores_long %>% filter(Score_Type != "TumorPurity")
#   purity_plot <- scores_long %>% filter(Score_Type == "TumorPurity")
#   # purity_plot <- arrange(purity_plot, Score)
#   
#   # Bar plot of all three scores
#   p1 <- ggplot(scores_plot, aes(x = reorder(Sample, Score), y = Score, fill = Score_Type)) +
#     geom_bar(stat = "identity", position = "dodge") +
#     labs(title = "ESTIMATE Scores by Sample",
#          x = "Sample",
#          y = "Score") +
#     theme_minimal() +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#     scale_fill_brewer(palette = "Set2")
#   
#   print(p1)
#   
#   # Plot purity separately
#   p2 <- ggplot(purity_plot, aes(x = reorder(Sample, Score), y = Score, col = Cohort)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "Estimated Tumor Purity by Sample",
#          x = "Sample",
#          y = "Tumor Purity") +
#     ylim(0, 1) +
#     theme_minimal() +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
#   
#   print(p2)
#   
#   return(list(scores = p1, purity = p2))
# }
# 
# # Create plots
# plots <- plot_estimate_scores(estimate_scores)

# Histogram of purity obtained per sample
ggplot(estimate_scores, aes(x = reorder(Sample, TumorPurity), y = TumorPurity, fill = Cohort)) +
    # geom_bar(stat = "identity") +
    geom_col() +
    labs(title = "Estimated Tumor Purity by Sample",
         x = "Sample",
         y = "Tumor Purity") +
    ylim(0, 1) +
    # theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Upper and lower bounds for 95% PI (approximate from paper)
estimate_scores$Upper95 <- pmin(estimate_scores$TumorPurity + 0.1, 1)
estimate_scores$Lower95 <- pmax(estimate_scores$TumorPurity - 0.1, 0)
# # More sophisticated version with calculated prediction intervals
# estimate_scores$Residual <- estimate_scores$TumorPurity - estimate_scores$PredictedPurity # it's the same!
# residual_se <- sd(estimate_scores$Residual)
# z_value <- qnorm((1 + 0.95) / 2)
# pi_width <- z_value * residual_se
# estimate_scores$Upper95 <- pmin(estimate_scores$PredictedPurity + pi_width, 1)
# estimate_scores$Lower95 <- pmax(estimate_scores$PredictedPurity - pi_width, 0)

# Create the plot
p <- ggplot() +
  # 95% prediction interval (grey shaded area)
  geom_ribbon(data = estimate_scores,
              aes(x = ESTIMATEScore, ymin = Lower95, ymax = Upper95),
              fill = "grey80", alpha = 0.5) +
  
  # Prediction line (solid black)
  geom_line(data = estimate_scores,
            aes(x = ESTIMATEScore, y = TumorPurity),
            color = "black", size = 1) +
  
  # Upper 95% PI dashed line
  geom_line(data = estimate_scores,
            aes(x = ESTIMATEScore, y = Upper95),
            color = "grey50", linetype = "dashed", size = 0.8) +
  
  # Lower 95% PI dashed line
  geom_line(data = estimate_scores,
            aes(x = ESTIMATEScore, y = Lower95),
            color = "grey50", linetype = "dashed", size = 0.8) +
  
  # Sample points (blue circles)
  geom_point(data = estimate_scores,
             aes(x = ESTIMATEScore, y = TumorPurity, label=Sample),
             color = "blue", size = 2.5, alpha = 0.7) +
             # geom_text(aes(label=Sample), hjust=0, vjust=0, check_overlap = TRUE, colour = "blue") +
             # geom_text_repel(hjust="top", vjust="top") +
             
  # Labels and theme
  labs(title = "Tumor Purity vs ESTIMATE Score",
       x = "ESTIMATE Score",
       y = "Tumor Purity") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    panel.grid.major = element_line(color = "grey90", size = 0.3)
  )

print(p)

#  plot corrélation purity et CT_perc
lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}
# pvalue <- paste("pvalue =", lmp(lm(Purity ~ CT_perc, coldata)))
# r2 <- paste("R2 =", round(summary(lm(Purity ~ CT_perc, coldata))$r.squared, 2))
lm_eqn <- function(df){
    m <- lm(Purity ~ CT_perc, df);
    eq <- substitute(italic(r)^2~"="~r2*","~~italic(p-value)~"="~pval, 
         list(r2 = format(summary(m)$r.squared, digits = 2),
              pval = format(lmp(m), digits=2)))
    as.character(as.expression(eq));
}

ggplot(coldata, aes(x=CT_perc, y=Purity)) +
    geom_point() +
    geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
    geom_text(x = 25, y = 1, label = lm_eqn(coldata), parse = TRUE)

# PUREE score ???

```

### Déconvolution

```{r deconvolution}

# MCP-counter


```


### PCAs sur tous les protein-coding gènes : {.tabset}

#### axis 1/2

```{r PCA both all ax1-2 v1, fig.width=13, fig.height=10}

# rld <- rlog(dds, blind=TRUE) # too slow for that many samples
rld <- vst(dds, blind=TRUE)

myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
to_be_plotted <- c("SampleID", "Indication", "Arm", "Site", "Age", "OS", "PFS", "best_response", "Nber_Prior_chemo", "Cycles_NP137", "Cycles_carboplatin", "Cycles_paclitaxel", "Cycles_pembrolizumab", "Purity", "Batch", "QC_uniq_aligned")
nber_top_genes <- nrow(count_matrix)

create_myPCAplots <- function(df, column, PCx, PCy, nber_top_genes) {
  pcaData <- plotPCA(rld, intgroup=column, pcsToUse=c(PCx,PCy), returnData=TRUE, ntop=nber_top_genes) #, ntop=1000)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  percent_x <- paste0(percentVar[1], "%")
  percent_y <- paste0(percentVar[2], "%")
  x <- paste0('PC', PCx)
  y <- paste0('PC', PCy)
  if (column == 'SampleID') {
    pca_plot <- ggplot(pcaData, aes(.data[[x]], .data[[y]], color=group, label=rownames(df))) +
    geom_point(size=3) +
    ggtitle(column) +
    # theme(plot.title = element_text(hjust=0.95, vjust=0.15, margin = margin(t = -20, b = 10, r = 5), size=11),
    theme(plot.title = element_text(margin = margin(t = 10, b = -20), size=11), legend.position="none", axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), plot.margin = unit(c(1, 1, 1, 1), "mm")) +
    geom_text_repel(hjust="top", vjust="top") +
    geom_density2d(alpha=.5) +
    coord_fixed() 
    return(list("plot" = pca_plot, "PCx" = percent_x, "PCy" = percent_y))
  }
  else if (column %in% numeric_columns) {
    sc_color <- scale_colour_gradientn(colours = myPalette(100), limits=c(min(df[,which(colnames(df) == column)]), max(df[,which(colnames(df) == column)])))
    pca_plot <- ggplot(pcaData, aes(.data[[x]], .data[[y]], color=group)) +
    geom_point(size=3) +
    ggtitle(column) +
    # theme(plot.title = element_text(hjust=0.95, vjust=0.15, margin = margin(t = -20, b = 10, r = 5), size=11),
    theme(plot.title = element_text(margin = margin(t = 10, b = -20), size=11), legend.position="none", axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), plot.margin = unit(c(1, 1, 1, 1), "mm")) +
    geom_density2d(alpha=.5) +
    coord_fixed() +
    sc_color
    return(list("plot" = pca_plot, "PCx" = percent_x, "PCy" = percent_y))
  }
  else {
    pca_plot <- ggplot(pcaData, aes(.data[[x]], .data[[y]], color=group)) +
    geom_point(size=3) +
    ggtitle(column) +
    theme(plot.title = element_text(margin = margin(t = 10, b = -20), size=11), legend.position="none", axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), plot.margin = unit(c(1, 1, 1, 1), "mm")) +
    geom_density2d(alpha=.5) +
    coord_fixed()
    return(list("plot" = pca_plot, "PCx" = percent_x, "PCy" = percent_y))
  }
}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/3
```{r PCA both all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/4
```{r PCA both all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/5
```{r PCA both all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### scree plot
```{r PCA both all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

#### contributions of individual samples to PCs 1-6
```{r PCA both all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

#### top genes contributing to PC variance
```{r PCA both all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

#### Biplots 
```{r PCA both all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

#### Top variables associated to PC axis 
```{r PCA both all top variables}

# Function to assess variable contribution to PCs
assess_variable_contribution <- function(rld, pca, ntop = 500, variables = NULL) {
  
  # If variables not specified, use all from colData
  if (is.null(variables)) { variables <- colnames(colData(rld)) }
  # # Calculate PCA # already done above, so added it to stdin
  # rv <- rowVars(assay(rld))
  # select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
  # pca <- prcomp(t(assay(rld)[select, ]))
  # Get PC coordinates
  pca_data <- data.frame(pca$x[, 1:2])
  pca_data <- cbind(pca_data, colData(rld))
  # Calculate variance explained
  percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100
    # For each variable, calculate correlation/association with PCs
  results_list <- list()
  
  for (var in variables) {
    # print(var)
    # Skip if variable doesn't exist or has only one level
    if (!var %in% colnames(pca_data)) next
    var_data <- pca_data[[var]]
    # Check if variable is numeric or categorical
    if (is.numeric(var_data) && !is.factor(var_data)) {
      next  # we want only the centered and reduced variables (factors)
      # # For numeric: calculate correlation
      # cor_pc1 <- cor(pca_data$PC1, var_data, use = "complete.obs")
      # cor_pc2 <- cor(pca_data$PC2, var_data, use = "complete.obs")
      # # R-squared as measure of variance explained
      # r2_pc1 <- cor_pc1^2 * 100
      # r2_pc2 <- cor_pc2^2 * 100
    } else if (is.factor(var_data) && endsWith(var, '_cr')) {
      var_data <- as.numeric(as.vector(pca_data[[var]]))
      cor_pc1 <- cor(pca_data$PC1, var_data, use = "complete.obs")
      cor_pc2 <- cor(pca_data$PC2, var_data, use = "complete.obs")
      # R-squared as measure of variance explained
      r2_pc1 <- cor_pc1^2 * 100
      r2_pc2 <- cor_pc2^2 * 100
    }
    else {
      # For categorical: ANOVA / calculate eta-squared
      var_data <- as.factor(var_data)
      # Skip if only one level
      if (length(levels(var_data)) < 2) next
      # ANOVA for PC1
      aov_pc1 <- aov(pca_data$PC1 ~ var_data)
      ss_total_pc1 <- sum((pca_data$PC1 - mean(pca_data$PC1))^2)
      ss_var_pc1 <- sum(aov_pc1$residuals^2)
      r2_pc1 <- (1 - ss_var_pc1/ss_total_pc1) * 100
      # ANOVA for PC2
      aov_pc2 <- aov(pca_data$PC2 ~ var_data)
      ss_total_pc2 <- sum((pca_data$PC2 - mean(pca_data$PC2))^2)
      ss_var_pc2 <- sum(aov_pc2$residuals^2)
      r2_pc2 <- (1 - ss_var_pc2/ss_total_pc2) * 100
    }
    
    results_list[[var]] <- data.frame(
      variable = var,
      PC1_variance_explained = r2_pc1,
      PC2_variance_explained = r2_pc2
    )
  }
  
  # Combine results
  contribution_df <- do.call(rbind, results_list)
  rownames(contribution_df) <- NULL
  # Sort by PC1 contribution
  contribution_df <- contribution_df[order(contribution_df$PC1_variance_explained, 
                                          decreasing = TRUE), ]
  
  return(list(
    contribution = contribution_df,
    percentVar = percentVar,
    pca_data = pca_data
  ))
}

# Use the function
var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>


#### Top variables contribution with PLS 
```{r PCA both all top variables PLS}

# Function to assess variable importance using PLS
assess_variable_pls <- function(coldata, rld, ntop = 500, ncomp = 2) {
  
  # Get expression data
  rv <- rowVars(assay(rld))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
  X <- t(assay(rld)[select, ])  # Samples x Genes
  # Prepare Y matrix - convert categorical to dummy variables
  Y_list <- list()
  variable_names <- colnames(coldata)
  
  for (var in variable_names) {
    var_data <- coldata[[var]]
    # Skip if all NA
    if (all(is.na(var_data))) next
    # Skip if ID or equivalent # on ne peut pas prendre les classifications, qui sont déterminées directement sur les gènes présents dans le RNAseq (tourne en rond)
    if (var == "SampleID" | var == "Date_traitement") next
    if (is.numeric(var_data) && !is.factor(var_data)) {
      next  # we want only the centered and reduced variables (factors)
    } else if (is.factor(var_data) && endsWith(var, '_cr')) {
      # Numeric variable - use as is (but centered and scaled)
        Y_list[[var]] <- as.numeric(as.vector(var_data))
    } else {
      # Categorical - convert to dummy variables
      var_factor <- as.factor(var_data)
      if (length(levels(var_factor)) >= 2) {
        # Create dummy variables (one-hot encoding)
        dummy <- model.matrix(~ var_factor - 1)
        colnames(dummy) <- paste0(var, "_", levels(var_factor))
        Y_list[[var]] <- dummy
      }
    }
    if (sum(is.na(var_data)) > 0) {
      new_rows <- data.frame(matrix(NA, nrow = length(which(is.na(var_data))), ncol = ncol(Y_list[[var]])))
      rownames(new_rows) <- which(is.na(var_data))
      colnames(new_rows) <- colnames(Y_list[[var]])
      Y_list[[var]] <- rbind(Y_list[[var]], new_rows)
      Y_list[[var]] <- Y_list[[var]][order(as.numeric(rownames(Y_list[[var]]))), ]
    }
  }
  
  # Combine into Y matrix
  Y <- do.call(cbind, Y_list)
  
  # Remove rows with NA
  complete_cases <- complete.cases(X, Y)
  X_clean <- X[complete_cases, ]
  Y_clean <- Y[complete_cases, ]
  
  # Run PLS
  pls_result <- pls(X_clean, Y_clean, ncomp = ncomp, mode = "regression")
  
  # Calculate variable importance for each component
  importance_list <- list()
  
  for (comp in 1:ncomp) {
    # Get loadings for Y variables
    y_loadings <- pls_result$loadings$Y[, comp]
    
    # Calculate proportion of variance explained by each original variable
    var_importance <- data.frame(
      variable = character(),
      loading = numeric(),
      importance = numeric(),
      stringsAsFactors = FALSE
    )
    
    current_col <- 1
    for (var in variable_names) {
      if (var %in% names(Y_list)) {
        n_cols <- ncol(as.matrix(Y_list[[var]]))
        var_cols <- current_col:(current_col + n_cols - 1)
        
        # Sum of squared loadings for this variable
        var_loading <- sum(y_loadings[var_cols]^2)
        
        var_importance <- rbind(var_importance, 
                               data.frame(variable = var,
                                        loading = sqrt(var_loading),
                                        importance = var_loading * 100,
                                        stringsAsFactors = FALSE))
        
        current_col <- current_col + n_cols
      }
    }
    
    importance_list[[paste0("Comp", comp)]] <- var_importance
  }
  
  return(list(
    pls_result = pls_result,
    importance = importance_list,
    var_explained = pls_result$prop_expl_var$X,
    Y_matrix = Y_clean,
    X_matrix = X_clean
  ))
}

# Use the function
pls_results <- assess_variable_pls(coldata, rld, ntop = nber_top_genes, ncomp = 2)


# Plot variable importance for both first components
plot_pls_importance <- function(pls_results) {
  # Component 1
  imp1 <- pls_results$importance$Comp1
  # imp1 <- imp1[order(imp1$importance, decreasing = TRUE), ]
  imp1$component <- "Component 1"
    # Component 2
  imp2 <- pls_results$importance$Comp2
  # imp2 <- imp2[order(imp2$importance, decreasing = TRUE), ]
  imp2$component <- "Component 2"
  
  imp_combined <- rbind(imp1, imp2)
  
  ggplot(imp_combined, aes(x = reorder(variable, importance), y = importance, fill = component)) +
    geom_bar(stat = "identity", position = "dodge") +
    coord_flip() +
    scale_fill_manual(values = c("Component 1" = "steelblue", 
                                  "Component 2" = "coral")) +
    labs(title = "Variable Importance - PLS Components 1 and 2",
         x = "Variable",
         y = "Importance (%)",
         fill = "Component") +
    theme_minimal()+
    theme(legend.position = "top")
  
}

plot_pls_importance(pls_results)

```

<br/>
<br/>



### PCAs sur les gènes les plus variants : {.tabset}

#### choosing optimal number of top variable genes
```{r PCA both variants optimal ntop}

# Function to calculate variance explained by top N genes
calculate_variance_by_ngenes <- function(rld, gene_numbers = seq(100, 2000, by = 100)) {
  
  # Get all gene variances
  rv <- rowVars(assay(rld))
  results <- data.frame(
    n_genes = integer(),
    variance_explained = numeric()
  )
  
  for (n in gene_numbers) {
    # Select top n genes
    select <- order(rv, decreasing = TRUE)[seq_len(min(n, length(rv)))]
    # Run PCA on these genes
    pca <- prcomp(t(assay(rld)[select, ]))
    # Calculate total variance explained by first 2 PCs
    percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100
    total_var_pc1_pc2 <- sum(percentVar[1:2])
    # Or you can use variance explained by all PCs (should be close to 100%)
    # For selection, typically we care about PC1+PC2
    
    results <- rbind(results, data.frame(
      n_genes = n,
      variance_explained = total_var_pc1_pc2
    ))
  }
  return(results)
}

# Calculate for different gene numbers
gene_numbers <- c(seq(100, 1000, by = 100), seq(1500, 3000, by = 500))
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

On peut choisir 600 gènes, c'est là où la pente de la courbe commence à être moins forte (méthode elbow). 

<br/>
<br/>

#### axis 1/2

```{r PCA both variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 600

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/3
```{r PCA both variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/4
```{r PCA both variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/5
```{r PCA both variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### scree plot
```{r PCA both variants scree plot}

# Calculate PCA of most variant genes
rv <- rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

#### contributions of individual samples to PCs 1-6
```{r PCA both variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

#### top genes contributing to PC variance
```{r PCA both variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

#### Biplots 
```{r PCA both variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

#### Top variables contribution to PC variance 
```{r PCA both variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

#### Top variables contribution with PLS 
```{r PCA both variants top variables PLS}

pls_results <- assess_variable_pls(coldata, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>


### Hierarchical Clustering Heatmaps :  {.tabset}

#### Sur tous les protein-coding gènes

```{r heatmaps both all genes}

rld_mat <- assay(rld)

# rld_cor_p <- cor(rld_mat)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

# Autre méthode pour plotter les Hierarchical clustering, donne des résultats différents (https://github.com/hamidghaedi/RNA-seq-differential-expression) :
# sampleDists <- dist(t(rld_mat))
# sampleDistMatrix <- as.matrix(sampleDists)
# rownames(sampleDistMatrix) <- colnames(rld_mat)
# colnames(sampleDistMatrix) <- NULL
# colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
# pheatmap(sampleDistMatrix,
#           clustering_distance_rows=sampleDists,
#           clustering_distance_cols=sampleDists,
#           col=colors)

```

<br/>
<br/>

#### sur les gènes les plus variants

```{r heatmaps both variant genes}

rld_mat_500 <- assay(rld)[select, ]

# rld_cor_p <- cor(rld_mat_500)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```






## Analyses sur le jeu de données Endo

```{r data overview Endo}

coldata_Endo <- coldata[which(coldata$Indication == "Endo"),]
coldata_Endo <- subset( coldata_Endo, select = -c(Indication, HPV_all, HPV16) ) # colonne qui ne prend qu'une valeur et colonnes de Cervix
count_matrix_Endo <- count_matrix[, which(coldata$Indication == "Endo")]

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_Endo,
  colData = coldata_Endo,
  design = ~ Arm
)

dds <- DESeq(dds)

# checking the DESeq2 dispersion 
plotDispEsts(dds, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Arm'): Cooks' distance")
# Cook’s distance is a measure of how much a single sample is influencing the fitted coefficients for a gene, and a large value of Cook’s distance is intended to indicate an outlier count

```

### PCAs sur tous les protein-coding gènes : {.tabset}

#### axis 1/2

```{r PCA Endo all ax1-2 v1, fig.width=13, fig.height=10}

rld <- vst(dds, blind=TRUE)

nber_top_genes <- nrow(count_matrix_Endo)
to_be_plotted <- c("SampleID", "Arm", "Site", "Age", "OS", "PFS", "best_response", "Nber_Prior_chemo", "Cycles_NP137", "Cycles_carboplatin", "Cycles_paclitaxel", "Cycles_pembrolizumab", "Batch", "ecog", "QC_uniq_aligned", "Purity")

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/3
```{r PCA Endo all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/4
```{r PCA Endo all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/5
```{r PCA Endo all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### scree plot
```{r PCA Endo all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

#### contributions of individual samples to PCs 1-6
```{r PCA Endo all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

#### top genes contributing to PC variance
```{r PCA Endo all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

#### Biplots 
```{r PCA Endo all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

#### Top variables associated to PC axis 
```{r PCA Endo all top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>


#### Top variables contribution with PLS 
```{r PCA Endo all top variables PLS}

pls_results <- assess_variable_pls(coldata_Endo, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>



### PCAs sur les gènes les plus variants : {.tabset}

#### choosing optimal number of top variable genes
```{r PCA Endo variants optimal ntop}

# Calculate for different gene numbers
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

Bien que la courbe ait une forme particulière, on peut choisir 1000 gènes (méthode elbow). 

<br/>
<br/>

#### axis 1/2

```{r PCA Endo variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 1000

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/3
```{r PCA Endo variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/4
```{r PCA Endo variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/5
```{r PCA Endo variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Endo, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### scree plot
```{r PCA Endo variants scree plot}

# Calculate PCA of most variant genes
rv <- rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

#### contributions of individual samples to PCs 1-6
```{r PCA Endo variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

#### top genes contributing to PC variance
```{r PCA Endo variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

#### Biplots 
```{r PCA Endo variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

#### Top variables contribution to PC variance 
```{r PCA Endo variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

#### Top variables contribution with PLS 
```{r PCA Endo variants top variables PLS}

pls_results <- assess_variable_pls(coldata_Endo, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>


### Hierarchical Clustering Heatmaps :  {.tabset}

#### Sur tous les protein-coding gènes

```{r heatmaps Endo all genes}

rld_mat <- assay(rld)

# rld_cor_p <- cor(rld_mat)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

```

<br/>
<br/>

#### sur les gènes les plus variants

```{r heatmaps Endo variant genes}

rld_mat_500 <- assay(rld)[select, ]

# rld_cor_p <- cor(rld_mat_500)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```





## Analyses sur le jeu de données Cervix

```{r data overview Cervix}

coldata_Cervix <- coldata[which(coldata$Indication == "Cervix"),]
coldata_Cervix <- subset( coldata_Cervix, select = -c(Indication, Mut_Status_endo, Estrogen_endo, Progesterone_endo, HR) ) # colonne qui ne prend qu'une valeur et colonnes qui n'ont que des NA
count_matrix_Cervix <- count_matrix[, which(coldata$Indication == "Cervix")]

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_Cervix,
  colData = coldata_Cervix,
  design = ~ Arm
)

dds <- DESeq(dds)

# checking the DESeq2 dispersion 
plotDispEsts(dds, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Arm'): Cooks' distance")

```

### PCAs sur tous les protein-coding gènes : {.tabset}

#### axis 1/2

```{r PCA Cervix all ax1-2 v1, fig.width=13, fig.height=10}

rld <- vst(dds, blind=TRUE)

nber_top_genes <- nrow(count_matrix_Cervix)

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/3
```{r PCA Cervix all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/4
```{r PCA Cervix all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/5
```{r PCA Cervix all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### scree plot
```{r PCA Cervix all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

#### contributions of individual samples to PCs 1-6
```{r PCA Cervix all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

#### top genes contributing to PC variance
```{r PCA Cervix all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

#### Biplots 
```{r PCA Cervix all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

#### Top variables associated to PC axis 
```{r PCA Cervix all top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>


#### Top variables contribution with PLS 
```{r PCA Cervix all top variables PLS}

pls_results <- assess_variable_pls(coldata_Cervix, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>



### PCAs sur les gènes les plus variants : {.tabset}

#### choosing optimal number of top variable genes
```{r PCA Cervix variants optimal ntop}

# Calculate for different gene numbers
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

On peut choisir 800 gènes, c'est là où la pente de la courbe commence à être moins forte (méthode elbow). 

<br/>
<br/>

#### axis 1/2

```{r PCA Cervix variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 800

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/3
```{r PCA Cervix variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/4
```{r PCA Cervix variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### axis 1/5
```{r PCA Cervix variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_Cervix, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

#### scree plot
```{r PCA Cervix variants scree plot}

# Calculate PCA of most variant genes
rv <- rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

#### contributions of individual samples to PCs 1-6
```{r PCA Cervix variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

#### top genes contributing to PC variance
```{r PCA Cervix variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

#### Biplots 
```{r PCA Cervix variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

#### Top variables contribution to PC variance 
```{r PCA Cervix variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

#### Top variables contribution with PLS 
```{r PCA Cervix variants top variables PLS}

pls_results <- assess_variable_pls(coldata_Cervix, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>


### Hierarchical Clustering Heatmaps :  {.tabset}

#### Sur tous les protein-coding gènes

```{r heatmaps Cervix all genes}

rld_mat <- assay(rld)

# rld_cor_p <- cor(rld_mat)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

```

<br/>
<br/>

#### sur les gènes les plus variants

```{r heatmaps Cervix variant genes}

rld_mat_500 <- assay(rld)[select, ]

# rld_cor_p <- cor(rld_mat_500)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```





# Conclusions et remarques 

L'indication sépare bien les échantillons (premier axe de l'ACP), ce qui est logique vu qu'il ne s'agit pas du même cancer : à bien séparer pour les DEGs<br/>
- pour Endo : les variables spécifiques à l'Endo (Progesteron et Estrogen) sont les meilleures en PLS<br/>
- pour Cervix : l'ecog est le meilleur prédicteur en PLS, à valider avec les LRT<br/>
- pour les deux : le Site de prélèvement est haut en PLS, ce qui est logique vu qu'il ne s'agit pas du même tissu<br/>
<br/>
Le choix du design sera finalisé avec l'étude des LRT.<br/>
<br/>
Il ne semble pas y avoir d'échantillon outlier : on peut tous les conserver.


# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
