---
title: "GyNet DEGs - avec DESeq2"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Gynet/"
knitr::opts_knit$set(root.dir = path)

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}
# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("DEGreport")
library("apeglm")
library("ggrepel")
library("GGally")
library("gridExtra")
library("mixOmics")
library(fgsea)
library(clusterProfiler)

library("dplyr")
library("ggplot2")
library('gplots')
library("gridExtra")
library("factoextra")
library(msigdbr)
library("reshape2")
library("tidyr")
library("splines")
library("ComplexHeatmap")
library(circlize)
library(viridis)
library(conflicted)
# install.packages("estimate", repos="http://R-Forge.R-project.org")

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Objectif

Etudier les gènes différentiellement exprimés sur les données GyNet, en séparant par indication (endomètre et col de l'utérus).<br/> 

<br/>
<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./coldata_GyNet.csv', sep=",")
coldata$BMI <- coldata$Weight / (coldata$Height)^2
coldata$HR <- ifelse(
  coldata$Estrogen_endo == "0" & coldata$Progesterone_endo == "0",
  "neg",
  "pos"
)
coldata$best_response2 <- case_when(
  coldata$best_response == "SD" ~ "SD_PD",
  coldata$best_response == "PD" ~ "SD_PD", 
  coldata$best_response == "PR" ~ "CR_PR",
  coldata$best_response == "CR" ~ "CR_PR",
  TRUE ~ NA_character_  # for any unexpected values
)
# coldata <- subset( coldata, select = - c(xx, xx))

# scale and center the numeric columns
numeric_columns <- c('Age', 'Weight', 'Height', 'BMI', 'OS', 'PFS', 'Cycles_NP137', 'Cycles_carboplatin', 'Cycles_paclitaxel', 'Cycles_pembrolizumab', 'Purity', 'CT_perc', 'QC_duplication_perc', 'QC_alignedM_Salmon', 'QC_uniq_aligned')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
# convert to factors the non-numeric columns
factor_columns <- colnames(coldata) %!in% numeric_columns 
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)


# countData (gene counts, not transcript's)
counts_genes <- read.table("./results_rnaseq_NFcore/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
#counts_genes <- as.matrix(counts_genes)

# keeping only the protein-coding genes, as we used a protocol capturing only exons (??? à valider auprès de Laurie !)
gtf <- import('/home/eugenie-modolo/Documents/Reference_files/gencode.v49.chr_patch_hapl_scaff.annotation.gtf')
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[which(rownames(counts_genes) %in% keep_genes), ]

count_matrix <- as.matrix(count_matrix_pcg[, count_columns])

coldata <- data.frame(coldata, row.names = count_columns)

coldata_wo_rep <- subset( coldata, select = -c(Age, Height, Weight, BMI, OS, PFS, Cycles_NP137, Cycles_carboplatin, Cycles_paclitaxel, Cycles_pembrolizumab, Purity, CT_perc, QC_duplication_perc, QC_alignedM_Salmon, QC_uniq_aligned) )

for (col in colnames(coldata_wo_rep)) {
    if (endsWith(col, '_cr')) {
        coldata_wo_rep[[col]] <- as.numeric(as.vector(coldata_wo_rep[[col]]))
    }
}

# creating the Endo and Cervix data
coldata_Endo <- coldata_wo_rep[which(coldata_wo_rep$Indication == "Endo"),]
coldata_Endo <- subset( coldata_Endo, select = -c(Indication, HPV_all, HPV16) ) # colonne qui ne prend qu'une valeur et colonnes qui n'ont que des NA
coldata_Endo <- droplevels.data.frame(coldata_Endo)
count_matrix_Endo <- count_matrix[, which(coldata_wo_rep$Indication == "Endo")]

coldata_Cervix <- coldata_wo_rep[which(coldata_wo_rep$Indication == "Cervix"),]
coldata_Cervix <- subset( coldata_Cervix, select = -c(Indication, Mut_Status_endo, Estrogen_endo, Progesterone_endo, HR) ) # colonne qui ne prend qu'une valeur et colonnes qui n'ont que des NA
coldata_Cervix <- droplevels.data.frame(coldata_Cervix)
count_matrix_Cervix <- count_matrix[, which(coldata_wo_rep$Indication == "Cervix")]

# removing the missing/problematic data for LRT
coldata_Endo_Primary <- coldata_Endo[which(coldata_Endo$Site == "Primary" & !is.na(coldata_Endo$Estrogen_endo) & !is.na(coldata_Endo$Progesterone_endo) & !is.na(coldata_Endo$best_response)),]
count_matrix_Endo_Primary <- count_matrix_Endo[, which(coldata_Endo$Site == "Primary" & !is.na(coldata_Endo$Estrogen_endo) & !is.na(coldata_Endo$Progesterone_endo) & !is.na(coldata_Endo$best_response))]
coldata_Cervix_Primary <- coldata_Cervix[which(coldata_Cervix$Site == "Primary" & !is.na(coldata_Cervix$HPV_all) & !is.na(coldata_Cervix$HPV16)),]
count_matrix_Cervix_Primary <- count_matrix_Cervix[, which(coldata_Cervix$Site == "Primary" & !is.na(coldata_Cervix$HPV_all) & !is.na(coldata_Cervix$HPV16))] 

# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

```

Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix)))` gène(s) au nom dupliqué, sur un total de `r nrow(count_matrix)` protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(coldata) == colnames(count_matrix)) == length(rownames(coldata))`.<br/>


## Indication : Endo

### Sites de prélèvement primaire uniquement

Design 1 : ~ HR + Arm + QC_duplication_perc_cr + Batch + Purity_cr + best_response2 <br/>
Design 2 : ~ HR + Arm + QC_duplication_perc_cr + Batch + Purity_cr + OS_cr <br/>

```{r DEGs Endo primary}

# design_1 <- as.formula("~ HR + Arm + QC_duplication_perc_cr + Batch + Purity_cr + best_response2")
# design_2 <- as.formula("~ HR + Arm + QC_duplication_perc_cr + Batch + Purity_cr + OS_cr")

design_1 <- as.formula("~ QC_duplication_perc_cr + Batch + Purity_cr + OS_cr")
design_2 <- as.formula("~ Arm + QC_duplication_perc_cr + Batch + Purity_cr + OS_cr")
design_3 <- as.formula("~ HR + Arm + QC_duplication_perc_cr + Batch + Purity_cr + OS_cr")

designs <- list(
  design_1 = design_1,
  design_2 = design_2,
  design_3 = design_3
)

selected_genes_list <- list()
norm.count_list <- list()
res_list <- list()
for (i in 1:3) {
  design_name <- paste0("design_", i)
  dds <- DESeqDataSetFromMatrix(
    countData = count_matrix_Endo_Primary,
    colData = coldata_Endo_Primary,
    design = designs[[design_name]]
  )
  dds <- DESeq(dds)
  norm.count <- assay(vst(dds, blind=FALSE))
  norm.count_list[[design_name]] <- norm.count
  res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
  res_list[[design_name]] <- res
  resOrdered <- res[order(res$padj), ]
  selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
  selected_genes_list[[design_name]] <- selected_genes
}

venn(selected_genes_list)

for ( i in 1:3 ) {
  design_name <- paste0("design_", i)
  print(c(design_name, length(selected_genes_list[[design_name]])))
}

# Volcano plots without shrink
for ( i in 1:3 ) {
  design_name <- paste0("design_", i)
  with(res_list[[design_name]], plot(log2FoldChange, -log10(padj), pch=20, main=paste("Volcano plot Endo design", i), cex=0.5))
  with(subset(res_list[[design_name]], padj<.05 ), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
  with(subset(res_list[[design_name]], abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="grey", cex=0.5))
  with(subset(res_list[[design_name]], padj <.05 & abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="green", cex=0.5))
}

# heatmaps des différents designs, en colorant par quelques variables d'intérêt
arm_col <- rainbow(length(unique(coldata_Endo_Primary$Arm)))
names(arm_col) <- sort(unique(coldata_Endo_Primary$Arm))
hr_col <- c("cyan","black")
names(hr_col) <- c("pos","neg")
bresp_col <- c("green", "darkgreen", "grey","red")
names(bresp_col) <- c("CR", "PR", "SD", "PD")
column_ha = HeatmapAnnotation(
  Arm=coldata_Endo_Primary$Arm,
  HR=coldata_Endo_Primary$HR,
  Best_response=coldata_Endo_Primary$best_response,
  col=list(Arm=arm_col,HR=hr_col,Best_response=bresp_col
  ))

distcor <- function(m){ as.dist(1-cor(t(m))  ) }
for ( i in 1:3 ) { 
  design_name <- paste0("design_", i)
  DEGs_gene_symbols <- counts_genes$gene_name[match(selected_genes_list[[design_name]], rownames(counts_genes))]
  # matrix_heatmap <- norm.count_list[[design_name]][which(rownames(norm.count_list[[design_name]]) %in% selected_genes_list[[design_name]]),]
  # rownames(matrix_heatmap) <- make.names(counts_genes$gene_name[match(rownames(matrix_heatmap), rownames(counts_genes))], unique = T)
  genes_ENSG <- row.names(norm.count_list[[design_name]])[match(selected_genes_list[[design_name]], rownames(norm.count_list[[design_name]]), nomatch = 0)]
  matrix_heatmap <- norm.count_list[[design_name]][genes_ENSG,]
  rownames(matrix_heatmap) <- DEGs_gene_symbols
  draw(Heatmap(as.matrix(matrix_heatmap),
               clustering_distance_columns  = "pearson",
               clustering_method_columns = "ward.D2",
               clustering_distance_rows  = "pearson",
               clustering_method_rows = "ward.D2",
               top_annotation = column_ha))
  heatmap.2(as.matrix(matrix_heatmap), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))
}

# # Volcano plot shrinked
# with(resLFCordered, plot(log2FoldChange, -log10(padj), pch=20, main="Volcano plot design2 shrinked", cex=0.5))
# with(subset(resLFCordered, padj<.05 ), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
# with(subset(resLFCordered, abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="grey", cex=0.5))
# with(subset(resLFCordered, padj <.05 & abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="green", cex=0.5))

# Autre code Volcano plot : (https://ciri.gitbiopages.ens-lyon.fr/bibs/carine/rnaseq_analysis_training/Session_6.html)
# res_tableOE_df <- res_list[[design_name]] %>%
#      as_tibble(rownames = "gene_ID") 
# res_tableOE_df <- res_tableOE_df %>%
#      mutate(is_up   = ifelse(padj < 0.05 & log2FoldChange > 0, "yes","no"),
#             is_down = ifelse(padj < 0.05 & log2FoldChange < 0, "yes","no"))
# res_tableOE_df <- res_tableOE_df %>% 
#      mutate(threshold_OE = (!is.na(padj)) & padj < 0.05 & abs(log2FoldChange) >= 0.58)
# ggplot(res_tableOE_df) +
#      geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold_OE)) +
#      ggtitle(paste("Volcano plot Endo design", i)) +
#      xlab("log2 fold change") + 
#      ylab("-log10 adjusted p-value") +
#      theme_bw() +
#      theme(legend.position = "none",
#            plot.title = element_text(size = rel(1.5), hjust = 0.5),
#            axis.title = element_text(size = rel(1.25))) 

# Other interesting figure (to be coded) (https://github.com/hamidghaedi/RNA-seq-differential-expression#4-visualization)
# visualize gene count data, for the DEGs, by violin plot differentiating by our condition of interest.


# cross-validation
cv_samples_DEGs <- function(count_matrix, coldata, design, padj.cutoff=0.05, lfc.cutoff=0.58) { #, coef
  
  # set.seed(123)
  cv_results <- list()
  cv_results_LFC <- list()

  for (outlier in 1:ncol(count_matrix)) {
    col_ind <- 1:ncol(count_matrix)
    indexes <- col_ind[-outlier]
    
    tryCatch({
      # Train on training set
      dds <- DESeqDataSetFromMatrix(
      countData = count_matrix[, indexes],
        colData = coldata[indexes, ],
        design = as.formula(design)
      )
      # dds$Time <- relevel(dds$Time, ref = "pre")
      dds <- DESeq(dds, quiet = TRUE)
      
      res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)
      resOrdered <- res[order(res$padj), ]
      selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
      cv_results[[outlier]] <- selected_genes
      
      # resLFC <- lfcShrink(dds, coef=coef, type="apeglm", lfcThreshold=lfc.cutoff)
      # resLFCordered <- resLFC[order(resLFC$svalue), ]
      # selected_genesLFC <- rownames(resLFCordered[which(resLFCordered$svalue < padj.cutoff/10), ])
      # cv_results_LFC[[outlier]] <- selected_genesLFC
      
    })
    }
  result <- list('all' = cv_results, 'LFC' = cv_results_LFC)
  return(result)
}

DEGs_in_common <- list()
for (i in 1:3) {
  design_name <- paste0("design_", i)
  check_repro <- cv_samples_DEGs(count_matrix_Endo_Primary, coldata_Endo_Primary, designs[[design_name]])
  
  min_nb <- round(nrow(coldata_Endo_Primary) * 0.95)
  all_DEGs <- do.call(c, check_repro$all)
  number_by_DEG <- table(all_DEGs)
  DEGs_in_common_here <- list()
  for (DEG in rownames(number_by_DEG)) {
    if (number_by_DEG[[DEG]] > min_nb) {
      DEGs_in_common_here <- append(DEGs_in_common_here, DEG)
    }
  }
  DEGs_in_common_here <- do.call(c, DEGs_in_common_here)
  print(c(design_name, length(DEGs_in_common_here)))
  DEGs_gene_symbols_here <- counts_genes$gene_name[match(DEGs_in_common_here, rownames(counts_genes))]
  DEGs_in_common[[design_name]] <- DEGs_in_common_here
  
  dds <- DESeqDataSetFromMatrix(
    countData = count_matrix_Endo_Primary,
    colData = coldata_Endo_Primary,
    design = designs[[design_name]]
  )
  # dds$Time <- relevel(dds$Time, ref = "pre")
  dds <- DESeq(dds)
  norm_count <- assay(vst(dds, blind=FALSE))
  # res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)
  
  genes_ENSG <- row.names(norm_count)[match(DEGs_in_common_here, rownames(norm_count), nomatch = 0)]
  matrix_heatmap <- norm_count[genes_ENSG,]
  rownames(matrix_heatmap) <- DEGs_gene_symbols_here
  heatmap.2(as.matrix(matrix_heatmap), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))

}
venn(DEGs_in_common)

```


## Indication : Cervix

### Sites de prélèvement primaire uniquement

Design 1 : ~ HPV16 + Arm + QC_uniq_aligned_cr + Batch + Purity_cr + best_response2 <br/>
Design 2 : ~ HPV16 + Arm + QC_uniq_aligned_cr + Batch + Purity_cr + PFS_cr <br/>

```{r DEGs Cervix primary}

# design_1 <- as.formula("~ HPV16 + Arm + QC_uniq_aligned_cr + Batch + Purity_cr + best_response2")
# design_2 <- as.formula("~ HPV16 + Arm + QC_uniq_aligned_cr + Batch + Purity_cr + PFS_cr")

# coefs <- list(
#   design_1 = "best_response2_SD_PD_vs_CR_PR",
#   design_2 = "PFS_cr"
# )

design_1 <- as.formula("~ QC_uniq_aligned_cr + Batch + Purity_cr + PFS_cr")
design_2 <- as.formula("~ Arm + QC_uniq_aligned_cr + Batch + Purity_cr + PFS_cr")
design_3 <- as.formula("~ HPV16 + Arm + QC_uniq_aligned_cr + Batch + Purity_cr + PFS_cr")

designs <- list(
  design_1 = design_1,
  design_2 = design_2,
  design_3 = design_3
)

selected_genes_list <- list()
norm.count_list <- list()
res_list <- list()
for (i in 1:3) {
  design_name <- paste0("design_", i)
  dds <- DESeqDataSetFromMatrix(
    countData = count_matrix_Cervix_Primary,
    colData = coldata_Cervix_Primary,
    design = designs[[design_name]]
  )
  dds <- DESeq(dds)
  norm.count <- assay(vst(dds, blind=FALSE))
  norm.count_list[[design_name]] <- norm.count
  res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
  res_list[[design_name]] <- res
  resOrdered <- res[order(res$padj), ]
  selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
  selected_genes_list[[design_name]] <- selected_genes
}

venn(selected_genes_list)

for ( i in 1:3 ) {
  design_name <- paste0("design_", i)
  print(c(design_name, length(selected_genes_list[[design_name]])))
}

# Volcano plots without shrink
for ( i in 1:3 ) {
  design_name <- paste0("design_", i)
  with(res_list[[design_name]], plot(log2FoldChange, -log10(padj), pch=20, main=paste("Volcano plot Cervix design", i), cex=0.5))
  with(subset(res_list[[design_name]], padj<.05 ), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
  with(subset(res_list[[design_name]], abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="grey", cex=0.5))
  with(subset(res_list[[design_name]], padj <.05 & abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="green", cex=0.5))
}

# heatmaps des différents designs, en colorant par quelques variables d'intérêt
arm_col <- rainbow(length(unique(coldata_Cervix_Primary$Arm)))
names(arm_col) <- sort(unique(coldata_Cervix_Primary$Arm))
hr_col <- c("cyan","black")
names(hr_col) <- c("pos","neg")
bresp_col <- c("green", "darkgreen", "grey","red")
names(bresp_col) <- c("CR", "PR", "SD", "PD")
column_ha = HeatmapAnnotation(
  Arm=coldata_Cervix_Primary$Arm,
  HR=coldata_Cervix_Primary$HR,
  Best_response=coldata_Cervix_Primary$best_response,
  col=list(Arm=arm_col,HR=hr_col,Best_response=bresp_col
  ))

distcor <- function(m){ as.dist(1-cor(t(m))  ) }
for ( i in 1:3 ) { 
  design_name <- paste0("design_", i)
  DEGs_gene_symbols <- counts_genes$gene_name[match(selected_genes_list[[design_name]], rownames(counts_genes))]
  # matrix_heatmap <- norm.count_list[[design_name]][which(rownames(norm.count_list[[design_name]]) %in% selected_genes_list[[design_name]]),]
  # rownames(matrix_heatmap) <- make.names(counts_genes$gene_name[match(rownames(matrix_heatmap), rownames(counts_genes))], unique = T)
  genes_ENSG <- row.names(norm.count_list[[design_name]])[match(selected_genes_list[[design_name]], rownames(norm.count_list[[design_name]]), nomatch = 0)]
  matrix_heatmap <- norm.count_list[[design_name]][genes_ENSG,]
  rownames(matrix_heatmap) <- DEGs_gene_symbols
  draw(Heatmap(as.matrix(matrix_heatmap),
               clustering_distance_columns  = "pearson",
               clustering_method_columns = "ward.D2",
               clustering_distance_rows  = "pearson",
               clustering_method_rows = "ward.D2",
               top_annotation = column_ha))
  heatmap.2(as.matrix(matrix_heatmap), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))
}

```



# Conclusion et remarques (à modifier !!)

.<br/>


# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
